Quantum Hoare Type Theory

### 2020-11-06 ms-paper/ms-paper.tex
Add definitions of Hermitian and projection operators

>>>>

\subsection{A Crash Course on Subspace Semantics}

Main references: \cite{logicqm36,ying_duan_feng_ji_2009,aqhl2019,unruh2019}.

\citet{ying_duan_feng_ji_2009} present a systematic exposition of predicate transformer semantics for quantum programs utilizing the algebraic structure of the Birkhoff-von Neumann quantum logic by restricting to quantum predicates expressible using projection operators. In this section, we will review how projection operators defined over Hilbert space can be viewed as predicates that will help us define the proposition language used in our Quantum Hoare Type Theory. We will also define various logial operations on these predicates that will help us reason about quantum programs.

This restriction to projection operators, which are Hermitian operators with 0 or 1 as their eigenvalues, will let us reason about quantum programs in a fashion similar to classical Boolean predicates. Further, we note that projection operators in Hilbert space correspond one to one with closed subspaces of this space. Let us start with some basics.

\begin{definition}[Hermitian operator]
	A Hermitian operator, $A$, in a finite-dimensional complex Hilbert space, $\mathcal{H}$, is a self-adjoint operator in $\mathcal{H}$. It is equal to its complex conjugate: $A = A^\dagger$.
\end{definition}
Hermitian operators describe physical observables and their eigenvalues are real numbers. \citet{dhondt_quantum_2006} define a quantum predicate in $\mathcal{H}$ as a Hermitian operator with eigenvalues within the unit interval ($[0, 1]$).

\citet{ying_duan_feng_ji_2009} point out a problem with that approach that it is not always possible to define certain logical operations such as conjunction and disjunction on some of these quantum predicates as they do not for a lattice. These operations are necessary to be able to compositionally reason about quantum programs. They, in turn, restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors. It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space. Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that is used to define a logic for quantum mechanics. Similar to \cite{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.

\begin{definition}[Projection operator]
	Let $X$ be a closed subspace of $\mathcal{H}$ and a vector $\ket{\psi} \in \mathcal{H}$. Then, there uniquely exist $\ket{\psi_0} \in X$ and $\ket{\psi_1} \in X^\bot$ (complement of X) such that $\ket{\psi} = \ket{\psi_0} + \ket{\psi_1}$. The vector $\ket{\psi_0}$ is called the projection of $\ket{\psi}$ onto subspace $X$ and is equal to $\ket{\psi_0} = P_X \ket{\psi}$. The operator $P_X$ is called the projection operator (or projector) onto $X$.
\end{definition}
